The Probability of Convergent Evolution During Local Adaptation.

### 2014-07-03 patchy-selection-paper.tex
abstract

@@

\maketitle
\begin{center}
$^1$ Computational Biology and Bioinformatics \\ University of
Southern California, Los Angeles, CA
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ University of California -- Davis. Davis, CA, 95616
\end{center}

\section{Introduction}

@@

The convergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{Stern:13}. 
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetics changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \cite{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

%\citep{wright2013indirect}
One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{macnair1991evolution,schat1996identical,turner2010serpentine};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches.
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes
\citep{benson1933concealing, DiceBlossom1937}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is primarily dark on a number of black lava flows separated by
much lighter soil \citep{dice1940ecologic, DiceBlossom1937}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible from the dark pigmentation adaptation on different outcrops \citep{hoekstra2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism}
and in lizards \citep{rosenblum2010molecular}.

@@

\paragraph{Concluding thoughts}
The falling cost of population genomic sequencing means that we will soon have the
opportunity to study the interplay of adaptation with geography and ecology 
across many populations within a species. 
Our work suggests that even quite geographically close populations 
may be forced to adapt convergently when migration is geographically limited. 
Thus, systems where populations have been repeatedly subject to 
strong local selection pressures may offer the opportunity 
to study highly replicated convergent adaptation within a similar
genetic background \citep{Stern:13}.  
Such empricial work will also strongly inform our understanding of how gene flow 
may work to keep ecologically similar populations evolving in concert \citep{sexton2013genetic}.
Our results suggest that adaptation to shared environments is
certainly no guarantee of a shared genetic basis to adaptation, 
suggesting that rapid adaptation to a shared environment could potentially drive
speciation under some scenerios \citep[in a manner similar to ][]{Kondrashov:03}.

>>>>

% The Spatial Resolution of Convergent Adaptation
% Convergent Evolution in Patchy Landscapes
% Parallel Evolution in Patchy Landscapes

@@

\maketitle
\begin{center}
$^1$ Computational Biology and Bioinformatics \\ 
University of Southern California, Los Angeles, CA \\
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ 
University of California -- Davis, Davis, CA, 95616
\end{center}

%%%%%%%%%%%
\abstract{
When the same phenotype is found to be responding to selection in parallel in isolated parts of a species range,
it provides strong evidence as to the degree of constraint imposed on evolution by the underlying genetic architecture.
Migration can counfound such inferences by synchronizing adaptation across the range;
but the importance of such sharing of local adaptations is unclear.
This paper studies the balance between these two in continuous environments with patchy selection pressures,
such as when an animal's coloration is selected to match its background,
but the species range contains large areas of differently colored substrate.
The main questions answered are: 
How long does it take for a new, locally adapted allele to appear in a patch of habitat through new mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of patch separation?
How can we tell which has occurred?
These are answered by decomposing the migration--selection equilibrium
surrounding an already adapted patch
into families of migrant alleles,
in particular treating those rare families that reach new patches
as spatial branching processes.
The speed at which successful migrant alleles traverse regions where they are maladapted
determines the length of the haplotype shared by around shared, adaptive alleles,
which is compared to the length of the haplotype shared within a patch.
Cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, is discussed as an empirical example.

}

%%%%%%%%%%% %%%%%%%%%%% 
\section{Introduction}

@@

The convergent evolution of similar phenotypes 
in response to shared selection pressures is a testiment to the power of selection 
to repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic}. 
\plr{want to also cite \citep{zhen2012parallel,martin2013repeated}?}
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetics changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \cite{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

%\citep{wright2013indirect}
One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{macnair1991evolution,schat1996identical,turner2010serpentine};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches.
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes
\citep{benson1933concealing}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is on a number of black lava flows has much darker pelage than on intervening,
much lighter soil \citep{DiceBlossom1937,dice1940ecologic}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible for the dark pigmentation found on different outcrops \citep{hoekstra2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism}
and in lizards \citep{rosenblum2010molecular}.

@@

\paragraph{Concluding thoughts}
The falling cost of population genomic sequencing means that we will soon have the
opportunity to study the interplay of adaptation with geography and ecology 
across many populations within a species. 
Our work suggests that even quite geographically close populations 
may be forced to adapt convergently when migration is geographically limited. 
Thus, systems where populations have been repeatedly subject to 
strong local selection pressures may offer the opportunity 
to study highly replicated convergent adaptation within a similar
genetic background \citep{stern2013genetic}.  
Such empricial work will also strongly inform our understanding of how gene flow 
may work to keep ecologically similar populations evolving in concert \citep{sexton2013genetic}.
Our results suggest that adaptation to shared environments is
certainly no guarantee of a shared genetic basis to adaptation, 
suggesting that rapid adaptation to a shared environment could potentially drive
speciation under some scenerios \citep[in a manner similar to ][]{Kondrashov:03}.
